Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-08-22 paper/paper.tex
Outline and some listings

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,abbreviation,definition,proof,lemma,theorem,corollary},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=fixed,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother

@@

\pagestyle{plain} % turn on page numbers in llncs

@@

\title{Verifying a Tableau Prover for Multi-Level-Syllogistics}
\author{Lukas Stevens}
\orcid{0000-0003-0222-6858}
\affiliation{%
  \institution{Technical University of Munich}
  \department{Department of Informatics}
  \streetaddress{Boltzmannstr. 3}
  \city{Garching}
  \postcode{85748}
  \country{Germany}
}
\email{lukas.stevens@in.tum.de}

@@

\subsection{Related work}
TODO

@@

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.
We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the type of a total function from \lstinline!'a! to \lstinline!'b!.
The types for booleans, natural numbers, and in integers are \lstinline!bool!, \lstinline!nat!, and \lstinline!int!, respectively.
Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
Analogously, we use \lstinline!'a list! to describe lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
For optional values, Isabelle/HOL offers the type \lstinline!option! where a term \lstinline!opt :: 'a option! is either \lstinline!None! or \lstinline!Some a! with \lstinline!a :: 'a!.
Finally, we remark that \textbf{iff} is equivalent to \lstinline!$=$! on type \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.

\section{A Semantics for Sets\label{sec:semantics}}
TODO

>>>>

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=fixed,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother

@@

@@

\title{Towards a Verified Tableau Prover for Multi-Level-Syllogistics}
\author{Lukas Stevens}
\orcid{0000-0003-0222-6858}
\affiliation{%
  \institution{Technical University of Munich}
  \department{Department of Informatics}
  \streetaddress{Boltzmannstr. 3}
  \city{Garching}
  \postcode{85748}
  \country{Germany}
}
\email{lukas.stevens@in.tum.de}

@@

\subsection{Related work}
AFP entry~\cite{hybrid_logic_afp} and workshop paper in Isabelle 2022 workshop.

@@

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.
We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the type of a total function from \lstinline!'a! to \lstinline!'b!.
The types for booleans, natural numbers, and in integers are \lstinline!bool!, \lstinline!nat!, and \lstinline!int!, respectively.
Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
Analogously, we use \lstinline!'a list! to describe lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
For optional values, Isabelle/HOL offers the type \lstinline!option! where a term \lstinline!opt :: 'a option! is either \lstinline!None! or \lstinline!Some a! with \lstinline!a :: 'a!.
Finally, we remark that \textbf{iff} is equivalent to \lstinline!$=$! on type \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
TODO: Add arrows and bracket notation for Pure formulae
\section{A Semantics for Sets\label{sec:semantics}}
\begin{lstlisting}
datatype 'a fm =
  Atom 'a | Neg ('a fm)
| And ('a fm) ('a fm) | Or ('a fm) ('a fm)
\end{lstlisting}

% TODO: Think about introducing notation for single
\begin{lstlisting}
datatype 'a pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\sqcup_\text{s}$ 'a pset_term
| 'a pset_term $\sqcap_\text{s}$ 'a pset_term
| 'a pset_term $-_\text{s}$ 'a pset_term
\end{lstlisting}

\begin{lstlisting}
datatype 'a pset_atom =
  'a pset_term $\in_\text{s}$ 'a pset_term
| 'a pset_term $\approx_\text{s}$ 'a pset_term
\end{lstlisting}

\begin{lstlisting}
type_synonym 'a pset_fm = 'a pset_atom fm
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}

\begin{lstlisting}
definition vars_branch :: 'a branch $\Rightarrow$ 'a set 
  vars_branch b $\equiv$ $\bigcup$(vars_fm ` set b)
\end{lstlisting}

\begin{lstlisting}
fun I$_\text{st}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  I$_\text{st}$ v $\emptyset$
| I$_\text{st}$ v (Var x) = v x
| I$_\text{st}$ v (Single s) = vset {I$_\text{st}$ v s}
| I$_\text{st}$ v (s $\sqcup_\text{s}$ t) = I$_\text{st}$ v s $\sqcup$ I$_\text{st}$ v t 
| I$_\text{st}$ v (s $\sqcap_\text{s}$ t) = I$_\text{st}$ v s $\sqcap$ I$_\text{st}$ v t 
| I$_\text{st}$ v (s $-_\text{s}$ t) = I$_\text{st}$ v s $-$ I$_\text{st}$ v t 
\end{lstlisting}

\begin{lstlisting}
fun I$_\text{sa}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ V where
  I$_\text{sa}$ v (s $\in_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ v s $\in$ elts (I$_\text{st}$ v t)
| I$_\text{sa}$ v (s $\approx_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ v s = I$_\text{st}$ v t
\end{lstlisting}

\begin{lstlisting}
fun subterms :: 'a pset_term $\Rightarrow$ 'a pset_term set
where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\sqcup_\text{s}$ t) = {s $\sqcup_\text{s}$ t}
    $\cup$ subterms s $\cup$ subterms t
| subterms (s $\sqcap_\text{s}$ t) = {s $\sqcap_\text{s}$ t}
    $\cup$ subterms s $\cup$ subterms t
| subterms (s $-_\text{s}$ t) = {s $-\text{s}$ t}
    $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) = {Single t} $\cup$ subterms t
\end{lstlisting}

\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\in_\text{s}$ t) = subterms s $\cup$ subterms t
| subterms_atom (s $\approx_\text{s}$ t) = subterms s $\cup$ subterms t
\end{lstlisting}

\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}

\begin{lstlisting}
definition subterms_branch ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms_branch b $\equiv$ $\bigcup$(subterms_fm ` set b)
\end{lstlisting}

\section{A Tableau Calculus for MLSS}
Interesting: they forgot one rule.

\section{Completeness of the Calculus}
\begin{lstlisting}
definition lin_sat b $\equiv$
  $\forall$b'. lextends b' b $\rightarrow$ set b' $\subseteq$ set b
\end{lstlisting}

\begin{lstlisting}
definition sat b $\equiv$
  lin_sat b $\land$ ($\nexists$bs'. fextends bs' b)
\end{lstlisting}

\subsection{Building a Model: Realisation of a Branch}
Idea: keep this short. The formalisation matches the pen-and-paper proof quite closely.
Show the results in Realisation of an Open Branch.

\subsection{Formalising the Key Technical Lemma}
Only one quite trivial invariant needed

\section{An Abstract Specification of the Decision Procedure}
Think about using the notion of coherent (or is this wfbranch?)
\begin{figure*}
  \centering
\begin{lstlisting}
function mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b $\Longrightarrow$
  mlss_proc_branch = mlss_proc_branch ((SOME b'. lextends b' b $\land$ set b $\subset$ set (b' @ b)) @ b)
|  $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$ $\Longrightarrow$
   mlss_proc_branch b = ($\forall$b' $\in$ (SOME bs. fextends bs b). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True

definition mlss_proc :: 'a pset_fm $\Rightarrow$ bool where
  mlss_proc $\phi$ $\equiv$ mlss_proc_branch [$\phi$]
\end{lstlisting}
\caption{Definition of the function \lstinline|mlss_proc_branch| and \lstinline|mlss_proc|.}
\end{figure*}

\subsection{Termination}
Maybe interesting: technical lemma that allows to obtain the point at which a parameter was introduced.
\subsection{Soundness}
Interesting: The model changes because auf new parameters.
Maybe move this to the chapter on the calculus?
\subsection{Lifting the Completeness of the Calculus}
TODO

\section{Future Work}
Challenges in obtaining an executable specification: membership cycle test unnecessary, fast branch membership test, proof reconstruction.
Additional constructs such as finiteness and pairs.
Optimisations for the algorithm.
